Working from home and digital divides: resilience during the pandemic

### 2021-01-02 paper/v1/broadband.speed.covid_anonymised.tex
submission

\title{Working from home and digital divides: resilience in the time of a
pandemic}

@@

\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O. Vriga\affilnum{3}}

\affiliation{\affilnum{1}{Department of Incredible Research, University A, City A, Country A}\\\affilnum{2}{Department of Applied Things, University B, City B, Country B}\\\affilnum{3}{Very Important Stuff Committee, Institute C, City C, Country C}}

\corrauth{Corresponding author name, This is sample corresponding address.}

\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}

@@

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds during the spring 2020 lockdown, may
reinforce or redress the spatial and social dimensions of digital
divisions. We employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of both
occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand. We also consider
what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era, and broadband services and
infrastructure must be fit for purpose.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the capability to telecommute also have the ability to
maintain their employment whilst protecting their health. However, the
success of these arrangements has been dependent upon the first level
digital divide, which is associated with access to and quality of
internet connectivity. \citet{SALEMINK2017360} provides a systematic
review of the pre-pandemic, first level digital divide in infrastructure
quality between urban and rural areas in various advanced economies.
Rural areas, predictably, fare worse. Yet whether this variation in
infrastructure quality affects the spatial footprint of telecommuting
has not previously been investigated, in part because telecommuting has
not previously been a cause of concentrated demand and pressure on
internet services.

@@

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2019.png}
\caption{Speed tests over time, 2019 \label{test2019}}
\end{figure}

@@

Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
speeds are much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
only cluster to experience higher speeds in the morning peak compared to
the evening, and thus the only cluster where the temporal profile of
internet use is closer to what might have been expected pre-pandemic.
And yet even among the other clusters, the reliability of internet
services during the working day still varies considerably. Interpreting
this variation from the large spikes and dips shown on Figures
\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
the most reliable internet services. The ratio of standard deviation to
mean in both these clusters is lowest, and speeds are only about 2\%
slower in the morning than the evening. Mean speeds are also higher than
in any other cluster, excluding cluster \(2\), where measures of
reliability suggest poorer performance.

Thus, broadband services in clusters \(9\) and \(11\), home to over
twelve million people, performed the best during the study period, in
terms of both quality and reliability. In Figure \ref{UpClusterL},
cluster \(11\) shows more noticeable peaks and troughs, with the lowest
points occurring, on average, between \(6:00-7:00\) on Monday morning,
\(14:00-15:00\) on Friday, and \(16:00-17:00\) on Thursday, not the
workday peak. Furthermore, these slow periods offer better speeds than
the average hourly profile of cluster \(13\). Finally, ignoring the
smallest clusters in terms of population, that is clusters \(4\) and
\(5\), clusters \(3\), \(7\) and \(10\) also have relatively high mean
upload speeds. Clusters \(3\) and \(10\) pair high mean speeds with low
standard deviations relative to the mean speeds, suggesting reliability
and resilience, as well as quality broadband services. Cluster \(7\) has
a higher ratio of standard deviation to mean, but there is less
difference in average speeds between the morning and evening peaks than
in clusters \(3\) and \(10\).

@@

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may explain why those in cluster \(7\) are the second
least likely to be served by Virgin Media. It is also a demonstration of
the complexity of both experienced broadband upload speeds as captured
by time-series clustering, and the geography of first level digital
divides as a product of quality as well as availability. In comparison,
cluster \(1\), which our analysis suggests lacks broadband resilience,
contains five, mainly rural authorities, but they are closer to a
metropolitan area than authorities in cluster \(3\), although furthest
from London, perhaps because they are scattered around the country.

@@

Yet we need to consider the results for other variables in Table
\ref{aux} to better determine whether clusters \(1\) and \(13\), which
appear to suffer most from poor quality internet services, are also more
likely to have a low skilled workforce, less able to benefit from
telecommuting. Cluster \(1\) has the lowest proportion of educated
people, and the lowest earnings, despite recording the highest
proportion of managerial and professional jobs, and the second highest
proportion of tech jobs. Rural areas such as those in cluster \(1\) are
home to many retired people, which might explain these results or
perhaps, as these figures are relative, we could note that cluster \(1\)
also has a greater proportion of skilled trades than other clusters. It
seems likely that the first level digital divide reinforces other
inequities in cluster \(1\), whilst those in cluster \(13\) are more
likely to earn more and have a better education despite poor internet
services. Cluster \(13\) also has the most businesses per inhabitant,
but is somewhere in the middle in terms of job density. If this is an
indication of a high number of SMEs, it could explain the variable
internet quality, considering that small businesses have not been seen
as the most valuable customers for higher speed broadband services if
they are not located near residential customers \citep{ofcom2016}.

Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
there is more noise within the averages we use to measure internet
quality and reliability. Our results indicate average mean speeds, and
Table \ref{aux} shows that cluster \(6\) also falls towards the middle
of the clusters on many of the socioeconomic variables. It ranks third
or fourth out of the eleven in terms of the likelihood of having a
higher proportion of managerial, professional, and tech jobs, as well as
higher earnings, but is fourth from bottom for educational attainment.
The LADs in cluster \(6\) are also diverse, with few truly remote areas,
but urban areas throughout England, including Birmingham, Leeds,
Sheffield, twelve London Boroughs, and many suburban areas and smaller
cities like Oxford and Cambridge. The capital cities of the other UK
nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
suggesting perhaps that the lower level of reliability discussed in
Section \protect\hyperlink{sec:4.1}{4.1} may be due to increased demand,
e.g.~for telecommuting, when it is not supported by the most resilient
internet connections.

In contrast, Clusters \(9\) and \(11\) enjoy resilient internet
connections, but LADs in Cluster \(9\) are more likely to host highly
educated individuals with higher earnings than cluster \(11\), which has
a negative coefficient for the NVQ4 variable. Cluster \(11\) has fewer
businesses per inhabitant and the second highest job density and cluster
\(9\) the second lowest. These coefficients might indicate that
resilient broadband infrastructure generates higher returns for those in
cluster \(9\), where slightly more slowdown in the AM peak was detected
(see Table \ref{up.cluster.descr}). LADs in cluster \(9\) are also less
likely to be in the South and those in cluster \(11\) more likely,
although the coefficient for cluster \(9\) might be skewed by the
presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
different economic profile than England. Still, both clusters consist of
larger LADs in terms of population, including the city centre among four
of the ten districts of Greater Manchester, eight London Boroughs, four
of the seven constituent authorities of the West Midlands Combined
Authority, Nottingham, and both Portsmouth and Southampton. Among the
\(65\) LADs are also a number of other tightly bounded urban areas, such
as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas
at the centre of less confined districts like Burnley, the digital hub
of Milton Keynes, and Northampton. These urban locations are on the
right side of the first level digital divide, and the regression results
suggest that internet resilience is supporting a wide range of small and
large urban economies.

LADs in clusters \(2\), \(3\), \(7\), \(10\) are also on the right side
of the first level digital divide, experiencing higher than average mean
speeds and ranking high to middle on measures of service reliability.
However, only LADs in cluster \(2\) and \(10\) are more likely to have
highly skilled workers to take advantage of working from home
opportunities. Cluster \(2\) is comprised of just two LADs, with the
lowest job density of any cluster. Cluster \(10\) is comprised of four
LADs, including two peripheral suburban areas of Birmingham, the East
London Borough of Newham, and Dundee. Suburbs are considered the most
likely urban form in which telecommuters live \citep{e2018does}, and
Dundee has a reputation for tech startups \citep{technation2017}. LADs
in cluster \(7\), which include the suburbs South of Belfast, a suburban
district of Greater Manchester, the villages between York and Leeds, and
a couple of small West Midlands towns have the highest earnings, and
thus may also be making the most of their resilient internet. In
contrast, LADs in cluster \(3\) are associated more with lower skills
and have the highest proportion of individuals in machine operation
jobs. Arguably, LADs in clusters \(3\) benefit the least from their
resilient internet infrastructure in an era when working from home
became a vehicle for economic resilience.

Finally, clusters \(8\) and \(12\) consist of LADs enjoying close to
average mean upload speeds, but opposing patterns of internet
reliability during the morning peak. LADs in cluster \(8\) are likely to
host more individuals employed in tech and administrative occupations
than any other cluster, as well as many in managerial occupations,
whilst the opposite applies to skilled trade and leisure occupations.
These LADs are characterised by a very small positive likelihood of more
educated residents, but are not significantly likely to earn more than
other clusters. Including suburbs near Leicester, Cardiff, Birmingham,
the green belt area south of London straddling the M25 motorway, and
rural West Devon, the LADs in cluster \(8\) seem to likely to have the
skills and occupations that would benefit from quality internet
services, but suffer from poor internet resilience and reliability.

In comparison, LADs in cluster \(12\) are less likely to be able to
benefit from quality internet services if they had them, with fewer
individuals achieving NVQ4 or better and lower levels of occupations
that would benefit from homeworking. Yet LADs in this cluster benefit
from the second highest level of earnings, and have the largest stock of
businesses per inhabitant and the highest job density. This density of
businesses could be why cluster \(12\) is the only cluster with higher
speeds in the morning peak during the study period. If people are at
home, business premises might well have been abandoned. Cluster \(12\)
is home to \(1.5\) million people spread across twelve LADs, including
the City of Westminster, another London Borough, and two others outside
London, but also a LAD in Northern Ireland, two in Wales, Preston in the
Northwest, and Fenland in the East of England. This spatial diversity
demonstrates that the temporal clustering of internet resilience is not
necessarily spatially dependent, and digital divides do not necessarily
overlap with economic ones.

@@

This paper offers a new perspective on telecommuting from the viewpoint
of the complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods, a family
of unsupervised machine learning techniques which are rarely utilised in
geographical research. Fast, reliable internet connections are necessary
for the population to be able to work from home. Although not every
place hosts individuals in occupations which allow for telecommuting nor
with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that without good internet
connectivity, these places would struggle more to achieve economic
resilience in a period like the current pandemic in any case. Indeed,
our analysis demonstrated that the temporal profiles of twelve of our
thirteen clusters had slower upload speeds in the morning than in the
evening. The opposite is likely to have been the norm prior to the
pandemic, as the level of demand and bandwidth management is the most
common cause of temporal variation in experienced speeds, and why
evening download speeds, rather than daytime upload speeds, have been
used to benchmark the performance of internet services. Thus, the new
patterns can be taken as evidence of widespread telecommuting and other
daytime internet use which changed the temporal profile of internet
activity throughout the UK, not just in areas with more digital industry
or better skills.

Upload speeds have not previously been seen as integral to universal
service, considering there has never before been such extreme demand for
telecommuting and operations such as video calls. This may be why the
average upload speeds for the largest two clusters -- \(6\) and \(13\)
-- contain so many LADs that are centres of the knowledge economy, from
Oxford and Cambridge in the former, to areas like Reading in the latter,
which has the highest concentration of digital businesses in the country
\citep{technation2017}. Thus, whilst some areas suffer from an
intersection of digital and economic divides, such as those in cluster
\(1\), others found their digital infrastructure to be less reliable
when confronted with the sudden change in the timing and type of demand.
Still other LADs, such as Milton Keynes in cluster \(9\) were able to
benefit both from reliable internet connections and populations which
were familiar with working from home and could capitalise on their
digital infrastructure. Yet the nuanced picture we gained through our
analysis of the UK case study suggests that quality internet
connectivity would also enable other LADs to gain ground during the
pandemic despite lower skills or earnings currently, especially in a
future where telecommuting might be a more common means of accessing
work and broadband services must be fit for purpose.

Indeed, the long-term effects of such drastic changes in telecommuting
and attitudes towards working from home are difficult to predict.
Nevertheless, they span various aspects of economy and society: from
changes to transportation planning due to altered commuting patterns to
changes in land use and urban planning to accommodate people who work
from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
productivity and innovation changes to changes in agglomeration
externalities and the attraction of large cities \citep{econobs}.
Further research may be able to measure the economic resilience of the
different clusters of places discussed in this paper once this pandemic
is firmly past. However, our analysis demonstrates that the economic
resilience made possible by working from home cannot be understood
without considering the underpinning digital divides and cannot be
achieved without planning for how the levels of digital, social and
economic divides might intersect.

@@

\pagebreak

>>>>

\title{Working from home and digital divides: resilience during the pandemic}

@@

\author{}

\affiliation{}

@@

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds during the spring 2020 lockdown, may
reinforce or redress the spatial and social dimensions of digital
divisions. We employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of
education, occupation and online accessibility intersect to enable or
hinder the practice of telecommuting at a time of extreme demand. We
also consider what lessons can be learned from this time for a future
where telecommuting is likely to remain a more common means of accessing
work, at least in comparison to the pre-Covid era.

@@

Our understanding of telecommuting as a product of enabled occupations
can be described as a manifestation of the third level digital divide,
as those who are able to use digital technologies to work from home
benefit from a high rate of return on their use of the internet in terms
of autonomy, flexibility, and time saved from commuting
\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
These returns have been even greater during the Covid-19 crisis, when
those with the capability to telecommute also have the ability to
maintain their employment whilst protecting their health. However, the
success of these arrangements has been dependent upon the first level
digital divide, which is associated with access to and quality of
internet connectivity. \citet{SALEMINK2017360} provides a systematic
review of the pre-pandemic, first level digital divide in infrastructure
quality between urban and rural areas in various advanced economies.
Rural areas, predictably, fare worse, yet as \citet{blank2018local}
highlight, variation in individual internet uptake and use is a product
of more complex spatial and demographic characteristics than simple
rurality or urbanisation. Yet whether this variation in infrastructure
quality affects the spatial footprint of telecommuting has not
previously been investigated, in part because telecommuting has not
previously been a cause of concentrated demand and pressure on internet
services.

@@

\begin{figure}
\centering
\includegraphics[width=0.85\textwidth,height=0.5\textheight]{figures/time.var.plot2019.png}
\caption{Speed tests over time, 2019 \label{test2019}}
\end{figure}

@@

Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
speeds are much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
only cluster to experience higher speeds in the morning peak compared to
the evening, and thus the only cluster where the temporal profile of
internet use is closer to what might have been expected pre-pandemic.
Among the other clusters, however, the reliability of internet services
during the working day still varies considerably. Interpreting this
variation from the large spikes and dips shown on Figures
\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
the most reliable internet services. The ratio of standard deviation to
mean in both these clusters is lowest, and speeds are only about 2\%
slower in the morning than the evening. Mean speeds are also higher than
in any other cluster, excluding cluster \(2\), where measures of
reliability suggest poorer performance.

Thus, broadband services in clusters \(9\) and \(11\), home to over
twelve million people, performed the best during the study period, in
terms of both quality and reliability. In Figure \ref{UpClusterL},
cluster \(11\) shows more noticeable peaks and troughs, with the lowest
points occurring, on average, between \(6:00-7:00\) on Monday morning,
\(14:00-15:00\) on Friday, and \(16:00-17:00\) on Thursday, not the
workday peak. Furthermore, these slow periods offer better speeds than
the average hourly profile of cluster \(13\). Clusters \(3\), \(7\) and
\(10\) also have relatively high mean upload speeds. Clusters \(3\) and
\(10\) pair high mean speeds with low standard deviations relative to
the mean speeds, suggesting reliability and resilience, as well as
quality broadband services. Cluster \(7\) has a higher ratio of standard
deviation to mean, but there is less difference in average speeds
between the morning and evening peaks than in clusters \(3\) and \(10\).

@@

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may explain why those in cluster \(7\) are the second
least likely to be served by Virgin Media. It is also a demonstration of
the complexity of both experienced broadband upload speeds as captured
by time-series clustering, and the geography of first level digital
divides as a product of quality as well as availability. In comparison,
cluster \(1\), which our analysis suggests lacks broadband resilience,
contains five, mainly rural authorities, but they are closer to a
metropolitan area than authorities in cluster \(3\), although furthest
from London, perhaps because they are scattered around the country --
see Figure \ref{map.up.clusters}.

@@

Yet we need to consider the results for other variables in Table
\ref{aux} to better determine whether clusters \(1\) and \(13\), which
appear to suffer most from poor quality internet services, are also more
likely to have a low skilled workforce, less able to benefit from
telecommuting. Cluster \(1\) has the lowest proportion of educated
people, and the lowest earnings, despite recording the highest
proportion of managerial and professional jobs, and the second highest
proportion of tech jobs. Rural areas such as those in cluster \(1\) are
home to many older, retired people \citep{blank2018local}, which might
explain these results or perhaps, as these figures are relative, we
could note that cluster \(1\) also has a greater proportion of skilled
trades than other clusters. In any case, it appears that the first level
digital divide reinforces other inequities in cluster \(1\), where the
even slower than average morning upload speeds shown in Table
\ref{up.cluster.descr} suggest that internet users were more active
during the working day. Meanwhile, those in cluster \(13\) are more
likely to earn more and have a better education despite poor internet
services. Cluster \(13\) also has the most businesses per inhabitant,
but is somewhere in the middle in terms of job density. If this is an
indication of a high number of SMEs, it could explain the variable
internet quality, considering that small businesses have not been seen
as the most valuable customers for higher speed broadband services if
they are not located near residential customers \citep{ofcom2016}.

Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
there is more noise within the averages we use to measure internet
quality and reliability. Our results indicate average mean speeds, and
Table \ref{aux} shows that cluster \(6\) also falls towards the middle
of the clusters on many of the socioeconomic variables. It ranks third
or fourth out of the eleven in terms of the likelihood of having a
higher proportion of managerial, professional, and tech jobs, as well as
higher earnings, but is fourth from bottom for educational attainment.
The LADs in cluster \(6\) are also diverse, with few truly remote areas,
but urban areas throughout England, including Birmingham, Leeds,
Sheffield, twelve London Boroughs, and many suburban areas and smaller
cities like Oxford and Cambridge. The capital cities of the other UK
nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
suggesting perhaps that the lower level of reliability discussed in
Section \protect\hyperlink{sec:4.1}{4.1} may be due to increased demand,
e.g.~for telecommuting, despite lacking the most resilient internet
connections.

In contrast, Clusters \(9\) and \(11\) enjoy resilient internet
connections, but LADs in Cluster \(9\) are more likely to host highly
educated individuals with higher earnings than cluster \(11\), which has
a negative coefficient for the NVQ4 variable. Cluster \(11\) has fewer
businesses per inhabitant and the second highest job density and cluster
\(9\) the second lowest. These coefficients might indicate that
resilient broadband infrastructure generates higher returns for those in
cluster \(9\), where slightly more slowdown in the AM peak was detected
-- see Table \ref{up.cluster.descr}. LADs in cluster \(9\) are less
likely to be in the South and those in cluster \(11\) more likely,
although the coefficient for cluster \(9\) might be skewed by the
presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
different economic profile than England. Still, both clusters consist of
larger LADs in terms of population, including districts within five of
England's ten largest cities, and a number of other stand-alone urban
areas and large market towns -- see Appendix
\protect\hyperlink{appendix1}{1}. These urban locations are on the right
side of the first level digital divide, and the regression results
suggest that internet resilience is supporting a wide range of small and
large urban economies.

LADs in clusters \(2\), \(3\), \(7\), \(10\) are also on the right side
of the first level digital divide, experiencing higher than average mean
upload speeds and ranking high to middle on measures of service
reliability. LADs in clusters \(2\) and \(10\) are also more likely to
have highly skilled workers to take advantage of working from home
opportunities. Cluster \(2\) is comprised of just two LADs, with the
lowest job density of any cluster. Cluster \(10\) is comprised of four
LADs, including two peripheral suburban areas of Birmingham, the East
London Borough of Newham, and Dundee. Suburbs are considered the most
likely urban form in which telecommuters live \citep{e2018does}, and
Dundee has a reputation for tech startups \citep{technation2017}. LADs
in cluster \(7\), which include the suburbs South of Belfast, a suburban
district of Greater Manchester, and some other Midlands and North
Yorkshire towns and villages have the highest earnings, and thus may
also be making the most of their resilient internet. In contrast, LADs
in cluster \(3\) are associated more with lower skills and have the
highest proportion of individuals in machine operation jobs. Arguably,
LADs in clusters \(3\) benefit the least from their resilient internet
infrastructure in an era when working from home became a vehicle for
economic resilience.

Finally, clusters \(8\) and \(12\) consist of LADs enjoying close to
average mean upload speeds, but opposing patterns of internet
reliability during the morning peak. LADs in cluster \(8\) are likely to
host more individuals employed in tech and administrative occupations
than any other cluster, as well as many in managerial occupations,
whilst the opposite applies to skilled trade and leisure occupations.
These LADs are characterised by a very small positive likelihood of more
educated residents, but are not significantly likely to earn more than
other clusters. Including suburbs near Leicester, Cardiff, and
Birmingham and south of London, the LADs in cluster \(8\) seem to likely
to have the skills and occupations that would benefit from quality
internet services, but suffer from poor internet resilience and
reliability.

In comparison, LADs in cluster \(12\) are less likely to be able to
benefit from quality internet services if they had them, with fewer
individuals achieving NVQ4 or better and lower levels of occupations
that would benefit from homeworking. Yet LADs in this cluster benefit
from the second highest level of earnings, and have the largest stock of
businesses per inhabitant and the highest job density. This density of
businesses could be why cluster \(12\) is the only cluster with higher
speeds in the morning peak during the study period. If people are at
home, business premises might well have been abandoned. Cluster \(12\)
is home to \(1.5\) million people spread across twelve LADs, from London
to Wales as listed in Appendix \protect\hyperlink{appendix1}{1}. This
spatial diversity demonstrates that the temporal clustering of internet
resilience is not necessarily spatially dependent, and digital divides
do not necessarily overlap with economic ones.

@@

This paper offers a new perspective on telecommuting from the viewpoint
of the complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods, a family
of unsupervised machine learning techniques which are rarely utilised in
geographical research. Fast, reliable internet connections are necessary
for the population to be able to work from home. Although not every
place hosts individuals in occupations which allow for telecommuting nor
with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that places without good
internet connectivity will still struggle more than other places to
achieve economic resilience in a period like the current pandemic when
internet resilience is so vital. Indeed, our analysis demonstrated that
the temporal profiles of twelve of our thirteen clusters had slower
upload speeds in the morning than in the evening. The opposite is likely
to have been the norm prior to the pandemic, as level of demand and
bandwidth management is the most common cause of temporal variation in
experienced speeds, and why evening download speeds, rather than daytime
upload speeds, have been used to benchmark the performance of internet
services. Thus, the new patterns can be taken as evidence of widespread
telecommuting and other daytime internet use which changed the temporal
profile of internet activity throughout the UK, not just in areas with
more digital industry or better skills.

Upload speeds have not previously been seen as integral to universal
service, considering there has never before been such extreme demand for
telecommuting and operations such as video calls. This may be why the
average upload speeds for the largest two clusters -- \(6\) and \(13\)
-- contain so many LADs that are centres of the knowledge economy, from
Oxford and Cambridge in the former, to areas like Reading in the latter,
which has the highest concentration of digital businesses in the country
\citep{technation2017}. Thus, whilst some areas suffer from an
intersection of digital and economic divides, such as those in cluster
\(1\), other areas, including cluster \(8\) as well as \(6\) and \(13\)
also found their digital infrastructure to be less than reliable when
confronted with the sudden change in the timing and type of demand. In
contrast, LADs such as the digital hub, Milton Keynes, in cluster \(9\)
were able to benefit both from reliable internet connections and
populations which were familiar with working from home and could
capitalise on their digital infrastructure. Yet the nuanced picture we
gained through our analysis of the UK case study suggests that quality
internet connectivity may also have enabled other LADs, such as those in
cluster \(3\) or \(11\) to gain ground during the pandemic despite lower
skill levels.

Digital infrastructure which considers upload speeds and working day
reliability as well as availability are likely to be particularly
important in a future where telecommuting might be a more common means
of accessing work and broadband services must be fit for purpose,
although the long-term effects of such drastic changes in telecommuting
and attitudes towards working from home are difficult to predict.
Nevertheless, they span various aspects of economy and society: from
changes to transportation planning due to altered commuting patterns to
changes in land use and urban planning to accommodate people who work
from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
productivity and innovation changes to changes in agglomeration
externalities and the attraction of large cities \citep{econobs}.
Further research may be able to measure the economic resilience of the
different clusters of places discussed in this paper once this pandemic
is firmly past. However, our analysis demonstrates that the economic
resilience made possible by working from home cannot be understood
without considering the underpinning digital divides and cannot be
achieved without planning for how the levels of digital, social and
economic divides might intersect.

@@
